Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-07 numerical-ppm.tex
Fixed up PPM section somewhat.  Still needs more work.

The primary hydrodynamic method used in \enzo\ is based on the
piecewise parabolic method (PPM) of
\citet{1984JCoPh..54..174C}, which has been
modified for the study of cosmological fluid flows.  
\citet{1984JCoPh..54..174C} describe two variants on this method: a
Lagrangian Remap variant and a Direct Eulerian variant.  Both have
been adapted for \enzo, and the Lagrangian Remap 
variant is described in \citet{1995CoPhC..89..149B}.  Because of its
conservative formulation, however, the Direct Eulerian variant is
preferred for \enzo\ AMR simulations.  We will describe the Direct
Eulerian (PPM-DE) variant here.

We can examine the structure of Equations~(\ref{eq:mass})-(\ref{eq:total_energy})
to obtain a better understanding of the effect of each term.
The derivatives are first order in time
and provide us with the dependent variables that will be evaluated each
time step: $\rhob$, $\vecvb$, and $E$.

The second term on the left hand side of each relation
is due to our choice of Eulerian coordinates and corresponds to
advection, the flow of conserved quantities 
($\rhob$, $\rhob \vecvb$, and $\rhob E$) 
across the mesh.  These will be termed the transport terms.

The first term on the right hand side of Equations~(\ref{eq:momentum}) and
(\ref{eq:total_energy}) comes from our choice of comoving coordinates
(a similar term also appears in Equation~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  It does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here, as they do not constitute a major
source of error.

The remainder of the terms (due to pressure and gravity) are denoted as
source terms.  They drive instabilities and discontinuities in the flow.
We make these distinctions because each of the three
groups of terms have different numerical behaviour and are therefore
treated differently.  The source and transport steps will be solved in
a self-consistent fashion by the PPM scheme while the expansion terms,
which are spatially localized, are split off and solved in a separate step
(as are radiative cooling and heating, if present).

The remainder of the terms (transport and source) in the fluid equations
involve derivatives.  
Because we are interested in phenomena with no special geometry, we will
restrict discussion to Cartesian coordinates.
Once again we use the concept of operator splitting,
but now apply it spatially.   We can rewrite the one-dimensional (Eulerian) 
versions of Equations~(\ref{eq:mass})--(\ref{eq:total_energy}) without expansion terms
in conservative form as,
\begin{eqnarray}
\frac{\partial \rho}{\partial t}  + \frac{1}{a} \frac{\partial \rho v }{\partial x}    & = &  
     0 \label{eq:mass1d} \\
\frac{\partial \rho v}{\partial t}  + \frac{1}{a} \frac{\partial \rho v^2}{\partial x}   + 
      \frac{1}{a} \frac{\partial p}{ \partial x} & = & 
      \rho \frac{1}{a} \frac{\partial \phi}{ \partial x}  \label{eq:momentum1d}  \\
\frac{\partial \rho E}{\partial t}  + \frac{1}{a} \frac{\partial \rho v E}{\partial x}  & =  &
      \rho v \frac{1}{a} \frac{\partial \phi}{\partial x}. \label{eq:total_energy1d}
\end{eqnarray}

@@

\subsubsection{The piecewise parabolic method in one dimension} % 2.2

We now restrict ourselves to the solution of Equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), plus transport, in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primary variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c_s$ and
$u+c_s$, where $c_s$ is the sound speed).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem.  This is described in detail in
\citet{1984JCoPh..54..174C}, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

@@

We have used subscripts to indicate zone-centered ($j$)
and face-centered ($j+1/2$) quantities, while superscripts refer to
position in time.  The cell width is $\Delta x_j$.
Although they have been discretized
in space, the accuracy of the updates depend on how well we can compute
the fluxes into and out of the cell during $\Delta t$.  
This in turn depends on our ability to compute the time-averaged 
values of $p$, $\rho$ and $v$ at the cell interfaces, denoted here
by $\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm 1/2}$, and $\overline{v}_{j\pm 1/2}$.  We now
describe the steps required to compute these quantities.
%Note that these difference equations are conservative.

We first construct monotonic piecewise 
parabolic (third order) interpolations in one dimension
for each of $p$, $\rho$, and $v$  The pressure is determined from
Equation~(\ref{eq:eq_of_state}), the equation of state.
%We have dropped the subscript b in this and the following two sections because we will be referring solely to baryonic quantities.  
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is Equation~(1.4) of CW84 (in the radial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
%The zones edges in the mass coordinates are $m_{j-1/2}$ and $m_{j+1/2}$ for the left and right edges, respectively.  
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

We have reduced the problem to finding $q_{L,j}$ and $q_{R,j}$.  While this
is simple in principle, it is complicated somewhat by the requirement that
these values be of sufficient accuracy and that the resulting distribution
be monotonic.  That is, no new maxima or minima are introduced.
The resulting formulae are straightforward but complicated and are not
reproduced here, but see Equations 1.7 to 1.10 of \citet{1984JCoPh..54..174C}.
We also optionally allow steepening as described in this reference.

In the Lagrangian case, once we have the interpolation constants we compute the characteristic domain for each zone edge.  This is simply the farthest a sound wave could travel in order to reach the interface by the end of a timestep. The characteristic domain encompasses all of the information that can reach the (Lagrangian) zone edge during the current timestep.  This information would then be fed into the Riemann solver to find the flux at each Lagrangian cell boundary.

In the Eulerian approach, this is more complex because the Eulerian
zone boundary does not naturally correspond to one the three characteristic waves and so a more complicated approach is required.  First, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$ or $u+c$ (where $c$ is the sound speed in a cell):
\begin{eqnarray}
      q^{C+}_{j+1/2} & = & f_{j+1/2,L}(\Delta t C^n_j), \\
      q^{C-}_{j+1/2} & = & f_{j+1/2,R}(-\Delta t C^n_{j+1}), \nonumber
\end{eqnarray}
where $q^{C+}$ ($q^{C-}$) refers to the left (right) characteristic $C$.  The functions $f$ are just integrals
over the parabolic interpolation formula:
%
\begin{eqnarray}
f_{j+1/2,L}(y) & = &
       {1 \over y} \int^{x_{j+1/2}}_{x_{j+1/2}-y} q_j(x) dx, \\
f_{j+1/2,R}(y) & = &
       {1 \over y} \int^{x_{j+1/2}+y}_{x_{j+1/2}} q_{j+1}(x) dx. \nonumber
\end{eqnarray}
The linearized gas dynamics equations are then used to compute second-order correct left and right states that take into account the multiple wave families.  This process is described in \citet{1984JCoPh..54..174C} and we use their equations (3.6) and (3.7).

With these effective states, an approximation to the Riemann problem is found
with an iterative approach \citep[see][]{Woodward86}, producing estimates for 
$\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm1/2}$, and $\overline{v}_{j\pm 1/2}$ that are third order
accurate in space and second order accurate in time.  These are then
used to solve the difference Equations (\ref{eq:mass_diff})--(\ref{eq:energy_diff}) for $\rho^{n+1}$, $v^{n+1}$, and $E^{n+1}$.

@@

To overcome this, we also solve the internal energy equation:
\begin{equation}
 \frac{\partial e}{\partial t} 
           + \frac{1}{a} \vecv \cdot \grad e
%         = - \frac{3(\gamma -1)\dot{a}}{a} e
         = - 3 \frac{\dot{a}}{a} \frac{p}{\rho}
           - \frac{p}{a\rho} \div \vecv
\end{equation}
in comoving coordinates.  The structure is similar to the total energy
equation; the second term on the left hand side represents transport, while
the first term on the right is due to expansion of the coordinate
system.  
It is differenced (again, in Eulerian form without the expansion term) as,
%
\begin{equation}
\rho_j^{n+1} e_j^{n+1}  = 
       \rho_j^n e_j^n   +  \Delta t  \left(
            \frac{  \avp{\rho} \avp{v} \avp{e}  - \avm{\rho} \avm{v} \avm{e}} {\Delta x_j} \right)
            - \Delta t \ p_j^{n} \left( \frac{ \avp{v} - \avm{v}} {\Delta x_j} 
                      \right)
                  \label{eq:gasenergy_diff}
\end{equation}
%
Note that because of the structure of this equation, it is not in
flux-conservative form.  In particular, the pressure is evaluated at
the cell center.  Unfortunately, time-centering of this pressure has
proved difficult to do without generating large errors in the internal
energy and so we conservatively leave the pressure at the old time in
this difference equation.  This leads to spreading of shocks; however,
we note that this equation is only used in hypersonic flows.

@@

It should be stressed that as long as the parameter ${\eta}_1$ is small enough
the dual energy method {\it will have no dynamical effect}.  We use
${\eta}_1 = 10^{-3}$, which is consistent with the truncation error of the
scheme for grid sizes that are typically used in our simulations.  We are now free to select the method by which the internal energy
field variable $e$ is updated so that it will not become contaminated with
errors advected by the total energy formulation but still give the
correct distribution in shocked regions. 
Since we are concerned with the advection of errors, the selection
criterion must look at each cell's local neighbourhood.
In one dimension, this is done with,
\begin{equation}
e = \cases{ (E - \vecv^2/2),& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) > {\eta}_2$,\cr
            e,& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) < {\eta}_2$.\cr}
    \label{eq:dualselect}
\end{equation}
Thus, ${\eta}_2$ determines when the synchronization (of $e$ with $E$) occurs.
%is a measure of how much error we will allow to be passed
%from neighboring cells.  
Too high a value may mask relatively weak shocks, 
while spurious heating (via contamination) may occur if it is set too low.
After some experimentation,
we have chosen ${\eta}_2 = 0.08$, a somewhat conservative value.
This scheme is optional and is generally only required in large-scale 
cosmological simulations where the gas cools due to the expansion of the universe
but large bulk flows develop due to the formation of structure.

@@

\subsubsection{Riemann Solvers and Fallback}

While most numerical hydrodynamics schemes center around finding approximate
solutions to the hydrodynamics equations, the core of philosophy of a Godunov's
method \citep{Godunov1959} is to find the exact solution to an
approximation of the physical setup.  The approximate problem in this case is
the Riemann problem, wherein two volumes are separated by a thin membrane that
is instantaneously removed at time $t=0$.  The subsequent evolution has an exact 
analytic solution.  This solution is
described in detail in many texts on computational fluid dynamics
\citep[e.g.,][]{toro-1997}.   In brief, there are three waves that propagate
away from the initial discontinuity.  The central wave, characterized by a
density jump but not a pressure jump, is called the contact discontinuity.  The
waves traveling to the left and right of the contact discontinuity can be either
shocks, if characteristics converge on the wave front, or  rarefaction fans if
characteristics diverge.

@@

For the MHD equations, there are seven waves, instead of three.  This makes the
exact solution to the Riemann quite a bit more expensive.  Both the HLL and HLLC
approximations can be formulated for the MHD equations, and are employed in both
Dedner and CT solvers in \enzo.  The HLLD
solver includes two of the additional waves, the rotational discontinuities,
making it a five-wave solver.

On rare occasions, high order
solutions can cause negative densities or energies.  Both PPM and MUSCL solvers employ a Riemann
solver fallback mechanism \citep{Lemaster09}.  If a negative density is found at
a particular interface, the more diffusive HLL Riemann solver is used instead.

% GB: I commented the following out because it has been incorporated into the above
%
%PPM is an explicit, higher order-accurate version of
%Godunov's method for ideal gas dynamics with third order-accurate piecewise parabolic
%monotonic interpolation and a nonlinear Riemann solver for shock
%capturing.  It advances the hydrodynamic equations in the following
%steps:
%\begin{enumerate}
% \item Construct monotonic parabolic interpolation of cell average
% data, for each fluid quantity.
% \item Compute interface states by averaging the parabola for over the domain of dependence for
% each interface
% \item Use interface data to solve the Riemann problem.
% \item Difference the interface fluxes to update the cell average quantities.
%\end{enumerate}
%It does an excellent job of capturing strong shocks in at
%most two cells.  Multidimensional schemes are built up by directional
%splitting and produce a method that is formally second order-accurate
%in space and time which explicitly conserves mass, linear momentum, and energy.

%To use the above algorithm for cosmological simulations, the
%conservation laws for fluid mass, momentum and energy 
%density are written in comoving coordinates for a
%Friedman-Robertson-Walker space-time, as described previously 
%in Equations~\ref{enzoconserve} 
%through~\ref{enzoenergy}.  Both the conservation laws and
%the Riemann solver are modified to include gravity, which is solved
%using an adaptive particle-mesh (PM) technique (see Section~\ref{sec.ov.nbody}).  
%The terms due to cosmological expansion, as well as primordial chemistry and
%radiative heating and cooling, are solved in a separate step because they have
%different numerical behavior, and therefore must be treated differently to ensure
%stability.  Note that unlike the \zeus\ hydro scheme, PPM does not need to use 
%artificial viscosity to resolve shocks.

%The system of equations described above works well for systems with
%relatively low Mach numbers, as long as these systems are well resolved.  
%However, cosmology is replete with situations where there are bulk hypersonic
%flows.  In these situations, the ratio of kinetic to thermal energy can be
%very high -- in some situations up to $10^6 - 10^8$.  This implies that the
%thermal energy is an extremely tiny fraction of the kinetic energy, which can 
%cause numerical problems when one is interested in just the thermal energy of the gas, since
% Equation~\ref{enzoenergy} solves for the total energy.  In this system of
%equations, the
% thermal energy $E_{therm}$ is calculated as $E - E_{kin}$, 
%where E is the total specific energy as
%calculated in equation~\ref{enzoenergy} and $E_{kin}$ is the specific kinetic energy,
%$0.5 \vecv_b^2$.  
%In hypersonic flows $E$ and $E_{kin}$ are nearly the same, and any 
%number calculated as 
%the difference of these is going to be strongly affected by numerical error.  To 
%avoid this problem, \enzo\ also solves the internal energy equation in
%comoving coordinates:

%\begin{equation}
%\frac{\partial e}{\partial t} + \frac{1}{a} \vecv_b \cdot \grad e = 
%- \frac{3(\gamma-1)\dot{a}}{a}e - \frac{p}{a \rho} \div \vecv_b
%\label{enzointenergy}
%\end{equation}

%In this equation $e$ is the internal energy (defined as $P/\rho (\gamma-1)$ and the other terms are as described
%previously.  The code still conserves total energy ($E$) as well.  In order
%to maintain consistency, both equations are solved at all times in all cells,
%with the equation for the total energy (eqtn.~\ref{enzoenergy}) being used
%for hydrodynamics routines and the internal energy $e$ being used when 
%temperature is required.  When pressure is required for dynamic purposes,
%the total energy is used if the ratio of thermal energy to total energy
%is less than some threshold value $\eta$, and the internal energy is used
%for values of the ratio larger than $\eta$.  A typical value of this parameter
%is $10^{-3}$.  This \emph{dual energy formulation} ensures that the
%method produces the correct entropy jump at strong shocks and also
%yields accurate pressures and temperatures in cosmological hypersonic
%flows.

>>>>

One (purely) hydrodynamic method used in \enzo\ is closely based on the
piecewise parabolic method (PPM) of
\citet{1984JCoPh..54..174C}, which has been
modified for the study of cosmological fluid flows.  
\citet{1984JCoPh..54..174C} describe two variants on this method:
Lagrangian Remap and Direct Eulerian; here we use the Direct Eulerian
version which is better suited for AMR simulations.  The Lagrangian
Remap version has previously been adapted for cosmological use
as described in \citet{1995CoPhC..89..149B} and the implementation
we use here is closely based on that work.

%We can examine the structure of Equations~(\ref{eq:mass})-(\ref{eq:total_energy})
%to obtain a better understanding of the effect of each term.
%The derivatives are first order in time
%and provide us with the dependent variables that will be evaluated each
%timestep: $\rhob$, $\vecvb$, and $E$.
%
%The second term on the left hand side of each relation
%is due to our choice of Eulerian coordinates and corresponds to
%advection, the flow of conserved quantities 
%($\rhob$, $\rhob \vecvb$, and $\rhob E$) 
%across the mesh.  These will be termed the transport terms.

The first term on the right hand side of equations~(\ref{eq:momentum}) and
(\ref{eq:total_energy}) come from our choice of comoving coordinates
(a similar term also appears in equation~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  A similar term does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here, as they do not constitute a major
source of error.  
We solve these terms (for all hydro methods)
using the same technique -- they are split from the rest of the terms and
solved using an (implicit) time-centered method, which is straightforward
as there are no spatial gradients.

%
%The remainder of the right-hand side terms (due to pressure and gravity) are denoted as
%source terms.  We make these distinctions because the 
%source and transport steps will be solved in
%a self-consistent fashion by the PPM scheme while the expansion terms,
%which are spatially localized, are split off and solved in a separate step
%(as are radiative cooling, heating and other processes, if present).

The remainder of the terms in the fluid equations
involve derivatives.  
Because we are interested in phenomena with no special geometry, we will
restrict discussion to Cartesian coordinates.
Once again we use the concept of operator splitting,
but now apply it spatially.   We can rewrite the one-dimensional (Eulerian) 
versions of equations~(\ref{eq:mass})--(\ref{eq:total_energy}) without expansion terms
in conservative form as,
\begin{eqnarray}
\frac{\partial \rho}{\partial t}  + \frac{1}{a} \frac{\partial \rho v }{\partial x}    & = &  
     0 \label{eq:mass1d} \\
\frac{\partial \rho v}{\partial t}  + \frac{1}{a} \frac{\partial \rho v^2}{\partial x}   + 
      \frac{1}{a} \frac{\partial p}{ \partial x} & = & 
      \rho \frac{1}{a} \frac{\partial \phi}{ \partial x}  \label{eq:momentum1d}  \\
\frac{\partial \rho E}{\partial t}  + \frac{1}{a} \frac{\partial \rho v E}{\partial x}  & =  &
      \rho v \frac{1}{a} \frac{\partial \phi}{\partial x}. \label{eq:total_energy1d}
\end{eqnarray}

@@

%\subsubsection{The piecewise parabolic method in one dimension} % 2.2

We now restrict ourselves to the solution of equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primative variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c_s$ and
$u+c_s$, where $c_s$ is the sound speed).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem.  This is described in detail in
\citet{1984JCoPh..54..174C}, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

@@

We have used subscripts to indicate zone-centered ($j$)
and face-centered ($j+1/2$) quantities, while superscripts refer to
position in time.  The cell width is $\Delta x_j$.
Although they have been discretized
in space, the accuracy of the updates depend on how well we can compute
the fluxes into and out of the cell during $\Delta t$.  
This in turn depends on our ability to compute the time-averaged 
values of $p$, $\rho$ and $v$ at the cell interfaces, denoted here
by $\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm 1/2}$, and $\overline{v}_{j\pm 1/2}$.  We now
describe the steps required to compute these quantities.

We first construct monotonic piecewise 
parabolic (third order) interpolations in one dimension
for each of $p$, $\rho$, and $v$.  The pressure is determined from
equation~(\ref{eq:eq_of_state}), the equation of state.
%We have dropped the subscript b in this and the following two sections because we will be referring solely to baryonic quantities.  
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is Equation~(1.4) of CW84 (in the spatial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
%The zones edges in the mass coordinates are $m_{j-1/2}$ and $m_{j+1/2}$ for the left and right edges, respectively.  
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

We have reduced the problem to finding $q_{L,j}$ and $q_{R,j}$.  While this
is simple in principle, it is complicated somewhat by the requirement that
these values be of sufficient accuracy and that the resulting distribution
be monotonic.  That is, no new maxima or minima are introduced.
The resulting formulae are straightforward but complicated and are not
reproduced here, but see Equations 1.7 to 1.10 of \citet{1984JCoPh..54..174C}.
We also optionally allow steepening as described in that reference.

Once we have the reconstruction, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$ or $u+c$ (where $c$ is the sound speed in a cell):
\begin{eqnarray}
      q^{C+}_{j+1/2} & = &
          {1 \over \Delta t C^n_j} \int^{x_{j+1/2}}_{x_{j+1/2}-\Delta t C^n_j} q_j(x) dx, \\
      q^{C-}_{j+1/2} & = & 
         {1 \over -\Delta t C^n_{j+1}} \int^{x_{j+1/2}+\Delta t C^n_{j+1}}_{x_{j+1/2}} q_{j+1}(x) dx \nonumber
\end{eqnarray}
where $q^{C+}$ ($q^{C-}$) refers to the left (right) characteristic $C$.  The linearized gas dynamics equations are then used to compute second-order accurate left and right states that take into account the multiple wave families.  This process is described in \citet{1984JCoPh..54..174C} and we use their equations (3.6) and (3.7).

With these effective states, an approximation to the Riemann problem is found
with an iterative approach \citep[see][and section~\ref{sec.riemann}]{Woodward86}, producing estimates for 
$\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm1/2}$, and $\overline{v}_{j\pm 1/2}$ that are third order
accurate in space and second order accurate in time.  These are then
used to solve the difference Equations (\ref{eq:mass_diff})--(\ref{eq:energy_diff}) for $\rho^{n+1}$, $v^{n+1}$, and $E^{n+1}$.

@@

To overcome this, we also solve the internal energy equation:
\begin{equation}
 \frac{\partial e}{\partial t} 
           + \frac{1}{a} \vecv \cdot \grad e
%         = - \frac{3(\gamma -1)\dot{a}}{a} e
         = - 3 \frac{\dot{a}}{a} \frac{p}{\rho}
           - \frac{p}{a\rho} \div \vecv
\end{equation}
in comoving coordinates.  The structure is similar to the total energy
equation; the second term on the left hand side represents transport, while
the first term on the right is due to expansion of the coordinate
system.  
It is differenced (again, in Eulerian form without the expansion term) as,
%
\begin{equation}
\rho_j^{n+1} e_j^{n+1}  = 
       \rho_j^n e_j^n   +  \Delta t  \left(
            \frac{  \avp{\rho} \avp{v} \avp{e}  - \avm{\rho} \avm{v} \avm{e}} {\Delta x_j} \right)
            - \Delta t \ p_j^{n} \left( \frac{ \avp{v} - \avm{v}} {\Delta x_j} 
                      \right)
                  \label{eq:gasenergy_diff}
\end{equation}
%
Note that because of the structure of this equation, it is not in
flux-conservative form.  In particular, the pressure is evaluated at
the cell center.  Unfortunately, time-centering of this pressure has
proved difficult to do without generating large errors in the internal
energy and so we leave the pressure at the old time in
this difference equation.  This leads to some spreading of shocks; however,
we note that this equation is only used in hypersonic flows.

@@

It should be stressed that as long as the parameter ${\eta}_1$ is small enough
the dual energy method {\it will have no dynamical effect}.  We use
${\eta}_1 = 10^{-3}$, which is consistent with the truncation error of the
scheme for grid sizes that are typically used in our simulations.  We are now free to select the method by which the internal energy
field variable $e$ is updated so that it will not become contaminated with
errors advected by the total energy formulation but still give the
correct distribution in shocked regions. 
Since we are concerned with the advection of errors, the selection
criterion must look at each cell's local neighbourhood.
In one dimension, this is done with,
\begin{equation}
e = \cases{ (E - \vecv^2/2),& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) > {\eta}_2$,\cr
            e,& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) < {\eta}_2$.\cr}
    \label{eq:dualselect}
\end{equation}
Thus, ${\eta}_2$ determines when the synchronization (of $e$ with $E$) occurs.
%is a measure of how much error we will allow to be passed
%from neighboring cells.  
Too high a value may mask relatively weak shocks, 
while spurious heating (via contamination) may occur if it is set too low.
After some experimentation,
we have chosen ${\eta}_2 = 0.1$, a somewhat conservative value.
This scheme is optional and is generally only required in large-scale 
cosmological simulations where the gas cools due to the expansion of the universe
but large bulk flows develop due to the formation of structure.

@@

\subsubsection{Riemann Solvers and Fallback}
\label{sec.riemann}

While most numerical hydrodynamics schemes center around finding approximate
solutions to the hydrodynamics equations, the core philosophy of a Godunov
method \citep{Godunov1959} is to find the exact solution to an
approximation of the physical setup.  The approximate problem in this case is
the Riemann problem, wherein two volumes are separated by a thin membrane that
is instantaneously removed at time $t=0$.  The subsequent evolution has an exact 
analytic solution.  This solution is
described in detail in many texts on computational fluid dynamics
\citep[e.g.,][]{toro-1997}.   In brief, there are three waves that propagate
away from the initial discontinuity.  The central wave, characterized by a
density jump but not a pressure jump, is called the contact discontinuity.  The
waves traveling to the left and right of the contact discontinuity can be either
shocks, if characteristics converge on the wave front, or  rarefaction fans if
characteristics diverge.

@@

For the MHD equations, there are seven waves, instead of three.  This makes the
exact solution to the Riemann quite a bit more expensive.  Both the HLL and HLLC
approximations can be formulated for the MHD equations, and are employed in both
the Dedner and CT solvers in \enzo.  The HLLD
solver includes two of the additional waves, the rotational discontinuities,
making it a five-wave solver.

On rare occasions, high order
solutions can cause negative densities or energies.  Both PPM and MUSCL solvers employ a Riemann
solver fallback mechanism \citep{Lemaster09}.  If a negative density is found at
a particular interface, the more diffusive HLL Riemann solver is used to compute the fluxes associated with that cell.
